---
title: "Algal/microbial diversity"
author: "Peter Fischer"
date: "6 april 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages



```{r}
library(vegan)
library(tidyverse)
library(pheatmap)
```

## Obtain data

Load data and rework them a bit so vegan can work with it. We also add metadata about algal diversity between samples and the season where each sample was taken.

```{r}
setwd("C:/Users/Peter/OneDrive - WageningenUR/fisch027/RData/Algae_paper")
data <- read.table("data/rel_abundance-PETERPCv2.txt",header=T)
row.names(data) <- data$Genus
data <- data[,-1]
data <- t(data)
Algal_population <- c("Mixed",rep(c("Chlorella dominated"),6),"Mixed","Mixed")
season <- c("Winter",rep(c("Spring"),3),rep(c("Summer"),3),"Autumn","Spring")
sampling <- row.names(data)

```


## Visualization of data

Now heatmaps in RStudio and markdown are a fickle thing, since graphs are limited to a certain size, which any decent heatmap always exceeds. Scaling down also isn't an option. What I generally do is fire up command line R in our server, then perform pheatmap there, where max plot size is much larger. You still might have to play with cell sizes though.


```{r}
pheatmap(t(data), cluster_cols = T,cluster_rows = F, cellwidth = 20,cellheight = 4)

```

## Visualization of selected data

The larger the data, the less readable a heatmap becomes. Thus, average each genus relative abundance, sort by most abundant genera, then pick the 50 most abundant and create a more readable heatmap with that data. 

```{r}
data1 <- read.table("data/rel_abundance-PETERPCv2.txt",header=T)

average <- rowMeans(data1[,2:10])
data1 <- cbind(data1,average)
data1 <- as_tibble(data1)
data1 <- arrange(data1, desc(average))
data1 <- as.data.frame(data1)
row.names(data1) <- data1$Genus
data1 <- data1[,-1]
data1 <- data1[,-10]
data1 <- data1[1:50,]

pheatmap(data1, cluster_cols = T,cluster_rows = F, cellwidth = 20,cellheight = 4)
```


## NMDS 

Here, we perform non-metric dimensional scaling on the bacterial population, inferring the influence of individual genera on the ordinal distribution of the sampling moments. A stressplot is created to check assumptions.

```{r }
nmds = metaMDS(data, distance = "bray")
stressplot(nmds)
plot(nmds)

```

## Visualization

Vegan plotting is very limiting, so we extract the data and build plots in ggplot2

```{r}
data.scores = as.data.frame(scores(nmds))


data.scores$Sampling <- sampling 
data.scores$'algal population' <- Algal_population
data.scores$Season <- season


gg = ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + geom_point(aes(colour=season,shape=`algal population`),size=5) + geom_text(aes(label=Sampling),hjust=0.45, vjust=-0.85) +
theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
legend.text = element_text(size = 12, face ="bold", colour ="black"), 
legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
legend.title = element_text(size = 14, colour = "black", face = "bold"), 
panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 0.2),
legend.key=element_blank()) + 
labs(x = "NMDS1", y = "NMDS2")  

gg  
```

## Statistics

We calculate the statistical significance of influence of algal diversity and season on the bacterial diversity between sampling moments using ANOSIM.  

```{r}
m_com = as.matrix(data)
ano_pop = anosim(m_com, Algal_population, distance = "bray", permutations = 9999)
ano_pop
ano_season = anosim(m_com, season, distance = "bray", permutations = 9999)
ano_season
```

## Influence of environmental parameters
First we need to load our metadata

```{r}
metadata <- read.table("data/metadata.txt",header=T)
row.names(metadata) <- metadata$Stage
metadata <- metadata[,-1]
metadata <- t(metadata)

### divide between environmental and biological parameters
environmentaldata <- c(1,2,3,4,5,9,10,11,12)
environmental <- metadata[,environmentaldata]

biologicaldata <- c(6,7,8,13,14,15,16,17,18)
biological <- metadata[,biologicaldata]
```

## Determine colinearity 

Colinearity negatively influence statistical interference, thus this should be addressed first. It also slims your list of explanatory variables making your plots more agreeable.  

```{r}
panel.cor <- function(x, y, digits=2, prefix="", cex.cor) # to create variable for colinearity plot
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y, use="complete.obs"))
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  txt <- paste(prefix, txt, sep="")
  if(missing(cex.cor)) cex <- 0.8/strwidth(txt)
  
  test <- cor.test(x,y)
  # borrowed from printCoefmat
  Signif <- symnum(test$p.value, corr = FALSE, na = FALSE,
                   cutpoints = c(0, 0.05, 0.1, 1),
                   symbols = c("*", ".", " "))
  
  text(0.5, 0.5, txt, cex = cex * r)
  text(.8, .8, Signif, cex=cex, col=2)
}
```

```{r}
pairs(biological, lower.panel = panel.smooth, upper.panel = panel.cor)
pairs(environmental,lower.panel=panel.smooth,upper.panel=panel.cor)
```

## RDA 
Redundancy analysis is performed to assess the influence of individual environmental parameters on the bacterial diversity. First, we perform Hellinger transformation to give weight to zero value data (see Legendre & Legendre, 2001).

Then, we perform RDA on the whole model to visualize the influence of parameters on ordination.

```{r}
hel.data <- decostand(data, "hellinger")

rdadata <- as.data.frame(hel.data)
rdaenv <- as.data.frame(environmental)
rdabio <- as.data.frame(biological)

var_rda <- rda(rdadata ~. , data= rdaenv, scale=T)
plot(var_rda,scaling=1,type="p",cex=1,main="all species RDA scores")
orditorp(var_rda,scaling=1,cex=1,display="sites",col="black")

var_rda <- rda(rdadata ~. , data= rdabio, scale=T)
plot(var_rda,scaling=1,type="p",cex=1,main="all species RDA scores")
orditorp(var_rda,scaling=1,cex=1,display="sites",col="black")

```

## Statistical influence of each individual parameter

This can also be done by a for loop, but for markdown viewing purposes it's easier to do it like this here. We perform anova on canonical axes (Legendre, Oksanen and ter Braak, 2011) 

```{r environmental}
var_rda_pars <- rda(hel.data ~ rdaenv[,1] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdaenv[,2] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdaenv[,3] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdaenv[,4] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdaenv[,5] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdaenv[,6] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdaenv[,7] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdaenv[,8] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdaenv[,9] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)
```

# biological parameters

```{r biological}

var_rda_pars <- rda(hel.data ~ rdabio[,1] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdabio[,2] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdabio[,3] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdabio[,4] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdabio[,5] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdabio[,6] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdabio[,7] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdabio[,8] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.data ~ rdabio[,9] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

```

### Algae data

We do the whole shebang again, this time with algal populations

```{r}

algaldata <- read.table("data/algae_abundance.txt",header=T)
row.names(algaldata) <- algaldata$Genus
algaldata <- algaldata[,-1]
algaldata <- t(algaldata)
season <- c("Winter",rep(c("Spring"),3),rep(c("Summer"),3),"Autumn","Spring")
sampling <- row.names(data)

```

## Visualization with heatmaps

```{r}
pheatmap(t(algaldata), cluster_cols = T,cluster_rows = F, cellwidth = 20,cellheight = 8)
```


## NMDS

```{r}
nmds = metaMDS(algaldata, distance = "bray")
stressplot(nmds)
plot(nmds)

data.scores = as.data.frame(scores(nmds))


data.scores$Sampling <- sampling 
data.scores$Season <- season


gg = ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + geom_point(aes(colour=season),size=5) + geom_text(aes(label=Sampling),hjust=0.45, vjust=-0.85) +
theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
legend.text = element_text(size = 12, face ="bold", colour ="black"), 
legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
legend.title = element_text(size = 14, colour = "black", face = "bold"), 
panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 0.2),
legend.key=element_blank()) + 
labs(x = "NMDS1", y = "NMDS2")  

gg  
```

## Anosim

Infer influence of season on diversity

```{r}
m_com = as.matrix(data)
ano_season = anosim(m_com, season, distance = "bray", permutations = 9999)
ano_season
```

## RDA

```{r }

hel.algaldata <- decostand(algaldata, "hellinger")
rdaalgaldata <- as.data.frame(hel.algaldata)

var_rda <- rda(rdaalgaldata ~. , data= rdaenv, scale=T)
plot(var_rda,scaling=1,type="p",cex=1,main="all species RDA scores")
orditorp(var_rda,scaling=1,cex=1,display="sites",col="black")

var_rda <- rda(rdaalgaldata ~. , data= rdabio, scale=T)
plot(var_rda,scaling=1,type="p",cex=1,main="all species RDA scores")
orditorp(var_rda,scaling=1,cex=1,display="sites",col="black")
```

## Statistics on parameter influence

```{r }
var_rda_pars <- rda(hel.algaldata ~ rdaenv[,1] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdaenv[,2] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdaenv[,3] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdaenv[,4] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdaenv[,5] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdaenv[,6] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdaenv[,7] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdaenv[,8] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdaenv[,9] , data=rdaenv, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)
```

# biological parameters

```{r }

var_rda_pars <- rda(hel.algaldata ~ rdabio[,1] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdabio[,2] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdabio[,3] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdabio[,4] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdabio[,5] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdabio[,6] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdabio[,7] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdabio[,8] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

var_rda_pars <- rda(hel.algaldata ~ rdabio[,9] , data=rdabio, scale=TRUE)
set.seed(111)
anova.cca(var_rda_pars, step=1000)

```
